---
title: "TFNC training 21 Feb 2023"
author: "Rie Goto"
date: "2023-02-18"
output:
  html_document: default
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Section 2: Working on Household Consumption and Expenditure Survey (HCES) data

## Part 1. Identify the dataset and variables to use in the household food consumption analysis

### 1.1 Pick up and discuss the data sets

HH_SEC_J1.csv - 60 food items\
HH_SEC_J3.csv - 10 food categories\
HH_SEC_J4.csv - share food with any non-HH members in the last 7 days\
HH_SEC_A.csv - HH location, data of interview\
HH_SEC_B.csv - individual id, age, sex, question 'did you eat meals in
this HH in the last 7 days?'\
HH_SEC_C.csv - education\
HH_SEC_F.csv - food consumption outside HH (Y/N)\
HH_SEC_M.csv - assets\
HH_SEC_V.csv - anthropometry (weight, height, MUAC)

First, let's focus on HH_SEC_J1

## Part 2. Data cleaning on HH_SEC_J1 - working on the food consumption data

### 2.1. import HH_SEC_J1.csv

```{r}
library(tidyverse)
library(here)
```

```{r}
dataJ1 <- read_csv(here("hces-data", "HH_SEC_J1.csv"))
```

### 2.2. change variable names

list of variable names and new names (in [ ])

-   sdd_hhid - household ID

-   itemcode - 60 food items

-   hh_j01 - Within the past 7 days, did the members of this household
    eat/drink any [...]? - YES/NO [cons_yn]

-   hh_j02_1 - if YES, how much in total did your household consume in
    the past 7 days? UNIT (kg, g, L, mL, pieces) [cons_unit]

-   hh_j02_2 - how much in total did your household consume in the past
    7 days? QUANTITY (numeric) [cons_quant]

-   hh_j03_1 - How much came from purchases during the past 7 days? UNIT
    [pur_unit]

-   hh_j03_2 - How much came from purchases during the past 7 days?
    QUANTITY [pur_quant]

-   hh_j04 - How much did you spend? (TSH) [pur_TSH]

-   hh_j05_1 - How much came from own-production? UNIT [prod_unit]

-   hh_j05_2 - How much came from own-production? QUANTITY [prod_quant]

-   hh_j06_1 - How much came from gifts and other sources? UNIT
    [gift_unit]

-   hh_j06_2 - How much came from gifts and other sources? QUANTITY
    [gift_quant]

### starting from a simple example

change variable names

Use %>% (and then do something) and rename

e.g. hces_subset_columns >- hces_data %\>% rename (food_sourse=gifted)

```{r}
dataJ1 <- dataJ1 %>%
  rename(cons_yn = hh_j01)
```

change other variables

```{r}
dataJ1 <- dataJ1 %>%
  rename(
    cons_unit = hh_j02_1,
    cons_quant = hh_j02_2,
    pur_unit = hh_j03_1,
    pur_quant = hh_j03_2,
    pur_THS = hh_j04,
    prod_unit = hh_j05_1,
    prod_quant = hh_j05_2,
    gift_unit = hh_j06_1,
    gift_quant = hh_j06_2
  )
```

check the data

```{r}
names(dataJ1)
str(dataJ1)
```

### 2.3. In pur_quant, prod_quant, gift_quant - there are ONE rather using NA. Therefore the variable type becomes character. Let's convert from NONE to NA.

Use mutate (modify variables) and na_if (change to NA if the condition
is...)

e.g. hces_subset_columns <- hces_data %>% mutate
(food_sourse=as.factor(food_sourse))

```{r}
dataJ1 <- dataJ1 %>%
  mutate(
    pur_quant = na_if(pur_quant, "NONE"),
    prod_quant = na_if(prod_quant, "NONE"),
    gift_quant = na_if(gift_quant, "NONE"),
  )
```

### 2.4. change variable types

```{r}
dataJ1 <- dataJ1 %>%
  mutate(
    itemcode = as.factor(itemcode),
    cons_yn = as.factor(cons_yn),
    cons_unit = as.factor(cons_unit),
    cons_quant = as.numeric(cons_quant),
    pur_unit = as.factor(pur_unit),
    pur_quant = as.numeric(pur_quant),
    prod_unit = as.factor(prod_unit),
    prod_quant = as.numeric(prod_quant),
    gift_unit = as.factor(gift_unit),
    gift_quant = as.numeric(gift_quant)
  )
```

check the data

```{r}
str(dataJ1)
```

### 2.5 counting frequency

1)  how many households had each food items (i.e. counting YES and NO)?

Use count

```{r}
food_freq <- count(dataJ1, itemcode, cons_yn)
```

One food item 'package/canned fish' was YES=0 and NO=1184, so omitted
YES (total obs. is 119, not 120). To keep YES in the counting, use
.drop=FALSE (do not drop any categories);

```{r}
food_freq <- count(dataJ1, itemcode, cons_yn, .drop = FALSE)
```

2)  frequency - select only Yes cases in cons_yn

Use filter to select only YES counting.

```{r}
food_freq_Y <- filter(food_freq, cons_yn == "YES")
```

3)  calculate and add percentage in food_freq_Y

Use mutate

```{r}
food_freq_Y <- food_freq_Y %>%
  mutate(percentage = n / 1184 * 100)
```

4)  adjust decimal places

Use round using 3 decimal places

```{r}
food_freq_Y <- food_freq_Y %>%
  mutate(percentage = round((n / 1184 * 100), 3))
```

### 2.6 unit conversion - check by food items

```{r}
food_unit <- count(dataJ1, itemcode, cons_unit)
```

select the food items using PIECES, or Litre and Millilitre - need to
discuss how to calculated them.

```{r}
count(dataJ1, itemcode, cons_unit)
```

```{r}
food_unit_P <- count(dataJ1, itemcode, cons_unit) %>% filter(cons_unit
== "PIECES")

food_unit_L <- count(dataJ1, itemcode, cons_unit) %>% filter(cons_unit
== "LITRE")

food_unit_mL <- count(dataJ1, itemcode, cons_unit) %>%
filter(cons_unit == "MILLILITRE")
```

### 2.7 Calculate food consumption in grams

Use if_else

```{r}
dataJ1 <- dataJ1 %>% mutate(cons_g=if_else(cons_unit=='GRAMS', cons_quant, 0))
```

Use case_when (when multiple conditions apply)

```{r}
dataJ1 <- dataJ1 %>% 
  mutate(cons_g=
           case_when(
             cons_unit=="GRAMS"~cons_quant,
             cons_unit=="KILOGRAMS"~(cons_quant*1000),
             cons_unit=="PIECES" & itemcode=="EGGS"~(cons_quant*70),
             TRUE~0
           ))
```

